A Non-Modal Analytical Method to Predict Turbulent Properties Applied to the Hasegawa-Wakatani Model

### 2014-12-26 HW_non-modal.tex
Dec 26

\title{A Non-Modal Analytical Procedure to Predict Turbulent Properties Applied to the Hasegawa-Wakatani Model}

@@

The calculation of linear instabilities through normal mode analysis, which solves for the eigenvalues and eigenvectors of a linearized dynamical system,
has successfully been used to predict turbulent properties such as turbulent onset and levels of transport for many plasma and fluid systems.
Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis has failed to predict turbulent onset in many experiments -- those in which the
turbulence is called subcritical~\cite{drazin1981}. 
In subcritical systems, turbulence develops even though all linear eigenvectors are stable such that infinitesimal perturbations on the equilibrium state cannot grow exponentially. 
Nevertheless, finite amplitude perturbations still excite turbulence through nonlinear instabilities.
Subcritical turbulence is common in plasmas as well, and several instances have been studied in the literature~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}.

@@

Moreover, even if normal mode fluctuations are unstable, they may not be as unstable as certain non-normal ones, which can produce turbulence with properties that are not predictable
from normal mode analysis.
When combined with nonlinear effects, non-modal fluctuations can sustain themsevles and have a permanent signature on turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
action of the linear operator. Transient growth, which can dominate turbulent evolution, must be found through non-modal calculations.

@@

To accomplish this, we model the effect that the nonlinearities have on the complex ratio of the state variables. Our model is one in which the 
nonlinearities randomize the complex ratios on a characteristic nonlinear time scale.
We estimate the nonlinear time scale by a characteristic linear time scale, taking either the inverse linear eigenmode frequency or growth rate as the linear time scale.
Between the nonlinear randomization, the system evolves linearly, and we use this evolution to calculate the non-modal growth rate. 
This non-modal growth rate takes the place of the eigenmode growth, which can be used
to give energy injection rates, turbulent saturation levels, and the turbulent onset criteria.

The paper is organized as follows: in Section~\ref{sec_hw_model}, we specify our illustrative model, the 2D and 3D Hasegawa-Wakatani model,
that we use throughout the paper and describe its linear non-normal properties,  while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
In Section~\ref{sec_nm_procedure}, we cover the details of our procedure to calculate the non-modal growth rate and compare it to growth rates derived directly from nonlinear numerical simulations. 
Finally, in the Appendix, we explore the possibility of predicting the critical gradient for turbulent onset with a modified subcritical 3D Hasegawa-Wakatani model.

@@

In order to illustrate and test our non-modal technique, we use as an example the well-known and well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}.
In this section, we analyze the linear normality of the model.
We consider both the original 2D and the extended 3D~\cite{biskamp1995} versions of the model in a straight, unsheared magnetic field. 
The HW model consists of two equations for the density $n$ and the electrostatic potential $\phi$. The 3D version of the model is

\beqar
\label{n_eq}
\pdiff{n}{t} = - {\mathbf v_E} \cdot \grad n - \kappa \pdiff{\phi}{y} + \xi \gradpar^2 (n - \phi) - D \gradperp^4 n, \\
\label{phi_eq}
\pdiff{\gradperp^2 \phi}{t} = - {\mathbf v_E} \cdot \grad (\gradperp^2 \phi) + \xi \gradpar^2 (n - \phi) - D \gradperp^6 \phi
\eeqar
The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $, where $\rho_s$ is the ion sound gyroradius at the electron temperature and $\omega_{ci}$ is
the ion cyclotron frequency. Additionally, $\mathbf{v_E} = \mathbf{b} \times \grad \phi$, where $\mathbf{b}$ is the magnetic field direction, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, where $\omega_{ce}$ is the electron cyclotron
frequency and $\nu_{ei}$ is the electron-ion collision frequency,
and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
energy on small scales. We set $D \to 10^{-5}$.
We associate the direction $\mathbf{b}$ as the $z$ direction. The direction perpendicular $\mathbf{b}$ that
points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these.

@@

The degree of non-normality of the linear operator in the HW model is largely a function of the adiabaticity parameter~\cite{camargo1998}, or in the 3D case, of $\xi \gradpar^2$. 
A normal matrix is on that commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$.
The HW model has a normal linear operator in the adiabatic limit: $\alpha, \xi \to \infty$. 
This is evident when we transform the linear operator into a linear matrix by using a Fourier decomposition. Because we use a zero-value boundary condition in the $x$ direction, the
Fourier decomposition takes the form:

@@

\beqar
\label{fourier_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) = \mathbf{A}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right), \\ \nonumber \\
\label{A_k}
\mathbf{A}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa k_y + \xi k_z^2 \\  \xi \frac{k_z^2}{k_\perp^2} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear operator, which we don't write explicitly.
Now in assessing the degree of normality of the HW model, we must make a choice of norms because one can always choose a norm that makes a particular matrix normal. However, it is much more revealing to choose
a physically motivated norm, especially the energy-norm, than one that simply makes a given matrix normal~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is

@@

\beqar
\label{fourier_en_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) = \mathbf{B}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right), \\ \nonumber \\
\label{B_k}
\mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} + \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
The square $L_2$-norm of the state vector $\frac{1}{\sqrt{2}} \left( n_k , k_\perp \phi_k \right)^T$ gives the energy contained in the mode $k$.
Therefore, we choose to cast the problem in terms of Eq.~\ref{fourier_en_eqn} rather than Eq.~\ref{fourier_eqn} and focus on the linear matrix $\mathbf{B}_k$ rather than $\mathbf{A}_k$. Thus, we can use
the well-known $L_2$-norm rather than a problem-dependent energy norm but still recover the same results as if we used an energy-norm with the formulation of Eq.~\ref{fourier_eqn}.

@@

\beq
\label{B_norm_limit0}
\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right).
\eeq
Thus, $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes. There are several ways to assess the non-normality of a matrix~\cite{trefethen2005}.
Perhaps the most intuitive is to use the scalar measure proposed by Henrici~\cite{henrici1962}:

\beq
\label{henrici_num}
H = \frac{||\mathbf{B} \mathbf{B}^\dagger - \mathbf{B}^\dagger \mathbf{B}||}{|| \mathbf{B} ||^2}
\eeq
where the operator $|| \cdot ||$ is the $L_2$-norm. This scalar normality measure can take on values: $0 \le H < 1$, where $H=0$ for a normal matrix. Since we are generally interested in highly non-normal
matrices where $H$ can be very close to $1$, we will look at the quantity $1-H$ rather than $H$ itself. Furthermore, in practice, what matters is how much transient energy growth is possible in the linear
system. The linear HW system is

\beq
\label{lin_HW}
\diff{u_k}{t} = \mathbf{B}_k u_k
\eeq
where $u_k = \frac{1}{\sqrt{2}} \left( n_k, k_\perp \phi_k \right)^T$.
The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u_k(0)$ where the exponential of the matrix is defined in terms of
its Taylor expansion. The solution is dependent on the initial condition $u_k(0)$.
The energy, which is synonymous with the energy amplification factor since we set $||u_k(0)||^2 =1$ evolves as

\beq
\label{Et}
E_k(t) = ||u_k(t)||^2 = ||e^{\mathbf{B}_k t} u_k(0)||^2
\eeq
Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such energy evolution curves is given by
$E_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on an initial condition.
Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
the fastest growing (or least damped) eigenmode of $\mathbf{B}_k$ (and of $\mathbf{A}_k$, since $\mathbf{B}_k$ is just a similarity transform of $\mathbf{A}_k$~\cite{camargo1998}). 
The equality holds when $\mathbf{B}_k$ is normal. 
Thus, non-normality in practice allows a system to undergo more energy growth than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that transient non-modal growth is strong for small values of the scalar normality measure $1-H$, which scales monotonically
with the adiabaticity parameter $\alpha$.

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
\caption{$E_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. Also indicated for each matrix is its scalar normality measure $1-H$.
The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
$||e^{\mathbf{B}_{k,\alpha} t}||^2$ displays large non-exponential growth at early times that increases with decreasing $\alpha$.}
\label{max_en_growth}
\end{figure}

Non-normality manifests itself in the eigenvectors -- non-normal matrices or operators have nonorthogonal linear eigenvectors. For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
with it. They are complex and take the form:

\beq
\label{lin_eigen_form}
\psi_{k,q} = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)_q {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} 
\eeq
where the subscript $q$ takes the values of $1$ and $2$, representing the two eigenvectors for each $k$. The two eigenvectors are different in that they have a different complex ratio $n_k/\phi_k$.
The two eigenvalues have different growth rates (the real part), but the same frequency, though with opposite sign.
As $\alpha, 1-H \to 0$, the two eigenvectors become less orthogonal -- more parallel or anti-parallel to one another. When the eigenvectors are nearly anti-parallel and initialized with finite amplitude,
their superposition, which constitutes the total system at $k$, may linearly evolve so that the norm of the superposition grows faster than either individual eigenvector. A paradigmatic illustration
of this may be seen in a review by Schmid~\cite{schmid2007}.

@@

\beqar
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}, \quad {\rm where}\\
\label{dEl_dt_def}
\diff{E_{l,k}}{t} = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) \\
\label{dEnl_dt_def}
 \diff{E_{nl,k}}{t} = \sum_{k'} T(k,k')
\eeqar
$\diff{E_{l,k}}{t}$ is composed of the linear terms on the RHS of Eq.~\ref{Explicit_En_eqn}, which are quadratic in the fluctuating quantities $n_k$ and $\phi_k$.
It describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion/viscosity.
$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. 
The terms in $T(k,k')$ are triadic in the fluctuating quantities and account for the energy exchange between fluctuations with different $k$'s. 
They are also conservative when summed over all wavenumbers: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$~\cite{camargo1995}.

@@

\beqar
\label{steady_state}
\gamma_{l,k} + \gamma_{nl,k} = \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt = 0, \\
\label{gamma_l_def}
{\rm where}, \ \gamma_{l,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt \\
\label{gamma_nl_def}
 \gamma_{nl,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt 
\eeqar
Thus, assuming steady or quasi-steady state turbulence $\gamma_{l,k} = - \gamma_{nl,k}$. Note that these, and all other growth rates, are defined to be time-independent (time-averaged) quantities.
At the scales of energy injection, $\gamma_{l,k}$ is positive and can be considered the turbulence analog of the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
In fact, $\gamma_{l,k} = \gamma_{s,k}$ if $n_k/\phi_k$ calculated from the nonlinear simulation is constant in time and equal to $n_k/\phi_k$ calculated from the fastest growing eigenmode at $k$. 
In general, though, these ratios are not the same because the turbulence at $k$ is composed of some linear combination of the \emph{two} eigenmodes.

\begin{table}
\begin{tabular}{|c|p {0.102\textwidth}|p {0.26\textwidth}|}
\hline
Symbol & Description & \qquad \qquad \quad Formula \\ \hline
$\gamma_s$ & Eigenmode growth rate & \qquad max(Re\{ eig( $\mathbf{B}_k$ )\})\\ \hline
$\gamma_l = \gamma_t$ & Turbulent growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt $\\ \hline
$\gamma_{nl}$ & Nonlinear growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt = - \gamma_{l}$\\ \hline
$\gamma_{{\rm nm}}$ & Non-modal growth rate & \ $\frac{1}{4 \tau_{nl,k}} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \tau_{nl,k}} e^{\mathbf{B}^{\dagger} \tau_{nl,k}} \} \right] $ \\ \hline
$\gamma_{\omega}$ & $\omega$ non-modal growth rate & \ $\frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right]$\\ \hline
$\gamma_{\gamma}$ & $\gamma$ non-modal growth rate & \ $\frac{\gamma_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \gamma_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \gamma_{s,k}^{-1}} \} \right]$\\ \hline
\end{tabular}
\caption{Growth rate symbol descriptions and formulas}
\label{gamma_table}
\end{table}

>>>>

\title{A Non-Modal Analytical Method to Predict Turbulent Properties Applied to the Hasegawa-Wakatani Model}

@@

The calculation of linear instabilities through normal mode analysis, which solves for the eigenvalues and eigenvectors of a linearized dynamical system,
has successfully been used to predict turbulent properties such as turbulent onset and levels of transport for many plasma and fluid systems.
Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis has failed to predict turbulent onset in many experiments -- those in which the
turbulence is called subcritical~\cite{drazin1981}. 
In subcritical systems, turbulence develops even though all linear eigenvectors are stable such that infinitesimal perturbations on the equilibrium state cannot grow exponentially. 
Nevertheless, finite amplitude perturbations excite turbulence through nonlinear instabilities.
Subcritical turbulence is common in plasmas as well, and several instances have been studied~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}.

@@

Moreover, even if normal mode fluctuations are unstable, they may not be as unstable as certain non-normal ones, which can produce turbulence with properties that are not predictable
from normal mode analysis.
When combined with nonlinear effects, non-modal fluctuations can sustain themselves and have a permanent signature on turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
action of the linear operator. Transient growth, which can dominate turbulent evolution, must be found through non-modal calculations.

@@

To accomplish this, we model the effect that the nonlinearities have on the complex ratio of the state variables. Our model is one in which the 
nonlinearities randomize the complex ratios on a characteristic nonlinear time scale.
We estimate the nonlinear time scale by a characteristic linear time scale, taking either the inverse linear eigenmode frequency or growth rate.
Between the nonlinear randomization, the system evolves linearly, and we use this evolution to calculate the non-modal growth rate. 
This non-modal growth rate takes the place of the eigenmode growth, which can be used
to give energy injection rates, turbulent saturation levels, and the turbulent onset criteria.

The paper is organized as follows: in Section~\ref{sec_hw_model}, we describe the linear non-normal properties of our illustrative model, the 2D and 3D Hasegawa-Wakatani model,
while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
In Section~\ref{sec_nm_procedure}, we cover the details of our procedure to calculate the non-modal growth rate and compare it to growth rates derived directly from nonlinear numerical simulations. 
We leave to the Appendix the unsuccessful attempt of predicting the critical gradient for turbulent onset with a modified subcritical 3D Hasegawa-Wakatani model.

@@

In order to illustrate and test our non-modal technique, we use as an example the well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}.
In this section, we analyze the linear normality of the model.
We consider both the original 2D and the extended 3D~\cite{biskamp1995} versions of the model in a straight, unsheared magnetic field. 
The HW model consists of two equations for the density $n$ and the electrostatic potential $\phi$. The 3D version of the model is

\beqar
\label{n_eq}
\pdiff{n}{t} = - {\mathbf v_E} \cdot \grad n - \kappa \pdiff{\phi}{y} + \xi \gradpar^2 (n - \phi) - D \gradperp^4 n, \\
\label{phi_eq}
\pdiff{\gradperp^2 \phi}{t} = - {\mathbf v_E} \cdot \grad (\gradperp^2 \phi) + \xi \gradpar^2 (n - \phi) - D \gradperp^6 \phi
\eeqar
The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , (x,y,z) \to (x,y,z)/\rho_s, t \to \omega_{ci} t $, where $\rho_s$ is the ion sound gyroradius at the electron temperature and $\omega_{ci}$ is
the ion cyclotron frequency. Additionally, $\mathbf{v_E} = \mathbf{b} \times \grad \phi$, where $\mathbf{b}$ is the magnetic field direction, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, where $\omega_{ce}$ is the electron cyclotron
frequency and $\nu_{ei}$ is the electron-ion collision frequency,
and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
energy on small scales. We set $D \to 10^{-5}$.
We associate the direction $\mathbf{b}$ as the $z$ direction. The direction perpendicular $\mathbf{b}$ that
points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these.

@@

The degree of non-normality of the linear operator in the HW model is largely a function of the adiabaticity parameter~\cite{camargo1998}, or in the 3D case, of $\xi \gradpar^2$. 
A normal matrix is one that commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$.
The HW model has a normal linear operator in the adiabatic limit: $\alpha, \xi \to \infty$. 
This is evident when we transform the linear operator into a linear matrix by using a Fourier decomposition. Because we use a zero-value boundary condition in the $x$ direction, the
Fourier decomposition takes the form:

@@

\beqar
\label{fourier_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) = \mathbf{A}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right), \\ \nonumber \\
\label{A_k}
\mathbf{A}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa k_y + \xi k_z^2 \\  \xi \frac{k_z^2}{k_\perp^2} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear operator, which we don't write explicitly.
Now in assessing the degree of non-normality of the HW model, we must make a choice of norms because one can always choose a norm that makes a particular matrix normal. However, it is much more revealing to choose
a physically motivated norm, especially the energy-norm, than one that simply makes a given matrix normal~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is

@@

\beqar
\label{fourier_en_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) = \mathbf{B}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right), \\ \nonumber \\
\label{B_k}
\mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} + \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
The square $L_2$-norm of the state vector $\frac{1}{\sqrt{2}} \left( n_k , k_\perp \phi_k \right)^T$ is the energy in the mode $k$.
Therefore, we choose to cast the problem in terms of Eq.~\ref{fourier_en_eqn} rather than Eq.~\ref{fourier_eqn} and focus on the linear matrix $\mathbf{B}_k$ rather than $\mathbf{A}_k$. Thus, we can use
the $L_2$-norm rather than a problem-dependent energy norm and still recover the same results as if we used an energy-norm with the formulation of Eq.~\ref{fourier_eqn}.

@@

\beq
\label{B_norm_limit0}
\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right).
\eeq
Thus, $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the non-normality of $\mathbf{B}_k$ changes. There are several ways to assess the non-normality of a matrix~\cite{trefethen2005}.
Perhaps the most intuitive is to use the scalar measure proposed by Henrici~\cite{henrici1962}:

\beq
\label{henrici_num}
H = \frac{||\mathbf{B} \mathbf{B}^\dagger - \mathbf{B}^\dagger \mathbf{B}||}{|| \mathbf{B} ||^2}
\eeq
where the operator $|| \cdot ||$ is the $L_2$-norm. This scalar non-normality measure can take on values: $0 \le H < 1$, where $H=0$ for a normal matrix. Since we are generally interested in highly non-normal
matrices where $H$ can be very close to $1$, we will look at the quantity $1-H$ rather than $H$ itself. Furthermore, in practice, what matters is how much transient energy growth is possible in the linear
system. The linear HW system is

\beq
\label{lin_HW}
\diff{u_k}{t} = \mathbf{B}_k u_k
\eeq
where $u_k = \frac{1}{\sqrt{2}} \left( n_k, k_\perp \phi_k \right)^T$.
The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u_k(0)$ where the exponential of the matrix is defined in terms of
its Taylor expansion. The solution is dependent on the initial condition $u_k(0)$.
The energy, which is synonymous with the energy amplification factor since we set $||u_k(0)||^2 =1$, evolves as

\beq
\label{Et}
E_k(t) = ||u_k(t)||^2 = ||e^{\mathbf{B}_k t} u_k(0)||^2
\eeq
Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such energy evolution curves is given by
$E_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on an initial condition.
Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
the fastest growing (or least damped) eigenmode of $\mathbf{B}_k$ (and of $\mathbf{A}_k$, since $\mathbf{B}_k$ is just a similarity transform of $\mathbf{A}_k$~\cite{camargo1998}). 
The equality holds when $\mathbf{B}_k$ is normal. 
Thus, non-normality in practice allows a system's energy to grow faster than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that transient non-modal growth is strong for small values of the scalar normality measure $1-H$, which scales monotonically
with the adiabaticity parameter $\alpha$.

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
\caption{$E_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. Also indicated for each matrix is its scalar non-normality measure $1-H$.
The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
$||e^{\mathbf{B}_{k,\alpha} t}||^2$ displays large non-exponential growth at early times that increases with decreasing $\alpha$.}
\label{max_en_growth}
\end{figure}

Non-normality manifests itself in the eigenvectors; non-normal matrices or operators have nonorthogonal linear eigenvectors. For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
with it. They are complex and take the form:

\beq
\label{lin_eigen_form}
\psi_{k,q} = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)_q {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} 
\eeq
where the subscript $q$ takes the values of $1$ and $2$, representing the two eigenvectors for each $k$. The two eigenvectors are different in that they have a different complex ratio $n_k/\phi_k$.
The two eigenvalues have different growth rates (the real part), but the same frequency, though with opposite sign.
As $\alpha, 1-H \to 0$, the two eigenvectors become less orthogonal -- more parallel or anti-parallel to one another. When the eigenvectors are nearly anti-parallel and initialized with finite amplitude,
their linear superposition constituting the component of the system at $k$ may linearly evolve so that the norm of the superposition grows faster than either individual eigenvector. A paradigmatic illustration
of this may be seen in a review by Schmid~\cite{schmid2007}.

@@

\beqar
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}, \quad {\rm where}\\
\label{dEl_dt_def}
\diff{E_{l,k}}{t} = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) \\
\label{dEnl_dt_def}
 \diff{E_{nl,k}}{t} = \sum_{k'} T(k,k')
\eeqar
$\diff{E_{l,k}}{t}$ is composed of the linear terms on the RHS of Eq.~\ref{Explicit_En_eqn}, which are quadratic in the fluctuating quantities $n_k$ and $\phi_k$.
It describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients minus the dissipation from collisions and hyperdiffusion/viscosity.
The terms in $T(k,k')$ are triadic in the fluctuating quantities and account for the energy exchange between fluctuations with different $k$'s. 
They are also conservative when summed over all wavenumbers: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$~\cite{camargo1995}.

@@

\beqar
\label{steady_state}
\gamma_{l,k} + \gamma_{nl,k} = \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt = 0, \\
\label{gamma_l_def}
{\rm where}, \ \gamma_{l,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt, \\
\label{gamma_nl_def}
 \gamma_{nl,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt 
\eeqar
Thus, assuming steady or quasi-steady state turbulence $\gamma_{l,k} = - \gamma_{nl,k}$. Note that these, and all other growth rates, are defined to be time-independent (time-averaged) quantities.
To keep track of all of the growth rates we define here and in the next section, refer to Table~\ref{gamma_table}.

At the scales of energy injection, $\gamma_{l,k}$ is positive and can be considered the turbulence analog of the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
In fact, $\gamma_{l,k} = \gamma_{s,k}$ if $n_k/\phi_k$ calculated from the nonlinear simulation is constant in time and equal to $n_k/\phi_k$ calculated from the fastest growing eigenmode at $k$. 
In general, though, these ratios are not the same because the turbulence at $k$ is composed of some linear combination of the \emph{two} eigenmodes, which have different $n_k/\phi_k$ ratios.

\begin{table}
\begin{tabular}{|c|p {0.102\textwidth}|p {0.26\textwidth}|}
\hline
Symbol & Description & \qquad \qquad \quad Formula \\ \hline
$\gamma_s$ & Eigenmode growth rate & \qquad max($Re\{ eig( \mathbf{B}_k)\}$)\\ \hline
$\gamma_l \equiv \gamma_t$ & Turbulent growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt $\\ \hline
$\gamma_{nl}$ & Nonlinear growth rate & \ \ \ $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt = - \gamma_{l}$\\ \hline
$\gamma_{{\rm nm}}$ & Non-modal growth rate & \ $\frac{1}{4 \tau_{nl,k}} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \tau_{nl,k}} e^{\mathbf{B}^{\dagger} \tau_{nl,k}} \} \right] $ \\ \hline
$\gamma_{\omega}$ & $\omega$ non-modal growth rate & \ $\frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right]$\\ \hline
$\gamma_{\gamma}$ & $\gamma$ non-modal growth rate & \ $\frac{\gamma_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \gamma_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \gamma_{s,k}^{-1}} \} \right]$\\ \hline
\end{tabular}
\caption{Growth rate symbols, descriptions, and formulas}
\label{gamma_table}
\end{table}
